\documentclass[10pt]{article}
\input{../templates/exam-note-template.tex}

\begin{document}

\begin{center}
{\LARGE Eigenvalue Quick Reference\par}
\end{center}
\notesubtitle{Concise revision notes for eigenvalue equations, checks, and basic matrix types.}
\source{Based on handwritten MIT 18.06 revision notes.}

\section{Core Idea}

\labelnote{Core idea}{Eigenvectors are directions that survive multiplication by \(A\). Along an eigenvector, the matrix only multiplies by a scalar.}

The defining equation is
\[
Ax=\lambda x,
\qquad x\ne0.
\]
Move all terms to one side:
\[
(A-\lambda I)x=0.
\]
For a nonzero solution, the matrix \(A-\lambda I\) must be singular:
\[
\det(A-\lambda I)=0.
\]

\section{Characteristic Equation}

For a \(2\times2\) matrix, the characteristic equation can be written as
\[
\lambda^2-\operatorname{trace}(A)\lambda+\det A=0.
\]
This is often the fastest way to check arithmetic.

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.26\linewidth}Y@{}}
\toprule
Quantity & Eigenvalue check \\
\midrule
Trace & Sum of eigenvalues. \\
Determinant & Product of eigenvalues. \\
Zero eigenvalue & Matrix is singular. \\
\bottomrule
\end{tabularx}
\end{center}

\section{Basic Examples}

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.32\linewidth}Y@{}}
\toprule
Matrix type & Eigenvalue memory \\
\midrule
Projection \(P\) with \(P^2=P\) & Eigenvalues are \(0\) or \(1\). \\
Permutation swapping two coordinates & Eigenvalues often include \(1\) and \(-1\). \\
Plane rotation by \(90^\circ\) & Eigenvalues are complex: \(i\) and \(-i\). \\
\bottomrule
\end{tabularx}
\end{center}

\section{Growth and Stability}

For repeated multiplication \(A^k\):
\[
|\lambda|<1
\]
means decay in that eigenvector direction.

\[
|\lambda|>1
\]
means growth in that direction.

\[
\lambda=1
\]
means that direction stays fixed.

\section{Common Mistakes}

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.40\linewidth}Y@{}}
\toprule
Mistake & Correct exam habit \\
\midrule
Allowing \(x=0\) as an eigenvector. & Eigenvectors must be nonzero. \\
Solving \(Ax=0\) for every eigenvector. & Solve \((A-\lambda I)x=0\). \\
Forgetting trace and determinant checks. & Use them after solving for \(\lambda\). \\
Assuming all rotations have real eigenvectors. & Plane rotations can have complex eigenvalues. \\
\bottomrule
\end{tabularx}
\end{center}

\section{Final Exam Checklist}

\tightlabel{Checklist}
\begin{enumerate}
  \item Write \(Ax=\lambda x\).
  \item Form \(A-\lambda I\).
  \item Solve \(\det(A-\lambda I)=0\).
  \item Check trace and determinant.
  \item For each \(\lambda\), solve \((A-\lambda I)x=0\).
  \item Use \(|\lambda|\) to judge growth or decay for powers.
\end{enumerate}

\end{document}
