\documentclass[10pt]{article}
\input{../templates/exam-note-template.tex}

\begin{document}

\begin{center}
{\LARGE 7.2 Singular Value Decomposition\par}
\end{center}
\notesubtitle{Exam notes for \(A=U\Sigma V^{T}\), singular values, and the four subspaces.}
\source{Based on handwritten MIT 18.06 revision notes.}

\section{Core Idea}

\labelnote{Core idea}{The SVD writes any matrix as orthogonal input directions, nonnegative stretching, and orthogonal output directions.}

For an \(m\times n\) matrix,
\[
A=U\Sigma V^{T}.
\]
Here \(V\) gives orthonormal directions in \(\mathbb{R}^n\), \(U\) gives orthonormal directions in \(\mathbb{R}^m\), and \(\Sigma\) contains the singular values.

\section{What the SVD Says}

\[
Av_i=\sigma_i u_i.
\]

This means: the right singular vector \(v_i\) is sent to the left singular vector \(u_i\), scaled by \(\sigma_i\ge 0\).

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.22\linewidth}Y Y@{}}
\toprule
Object & Lives in & Meaning \\
\midrule
\(v_i\) & \(\mathbb{R}^n\) & Orthonormal input direction. \\
\(u_i\) & \(\mathbb{R}^m\) & Orthonormal output direction. \\
\(\sigma_i\) & Nonnegative number & Stretch factor from \(v_i\) to \(u_i\). \\
\(\Sigma\) & \(m\times n\) diagonal shape & Stores singular values in decreasing order. \\
\bottomrule
\end{tabularx}
\end{center}

\section{How to Compute It}

The right singular vectors come from the symmetric matrix \(A^{T}A\):
\[
A^{T}A v_i=\sigma_i^2 v_i.
\]

So the singular values are
\[
\sigma_i=\sqrt{\lambda_i(A^{T}A)}.
\]

For each \(\sigma_i>0\), compute the matching left singular vector by
\[
u_i=\frac{Av_i}{\sigma_i}.
\]

If \(U\) or \(V\) is not complete, add orthonormal vectors to finish the basis.

\section{Geometry}

The SVD is the clean geometric order:
\[
x
\xrightarrow{\;V^{T}\;}
\text{rotate/reflection}
\xrightarrow{\;\Sigma\;}
\text{stretch}
\xrightarrow{\;U\;}
\text{rotate/reflection}.
\]

The singular values show the principal stretches of the matrix.

\section{Rank and the Four Subspaces}

If \(\sigma_1,\ldots,\sigma_r>0\) and the rest are zero, then \(r=\operatorname{rank}(A)\).

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.25\linewidth}Y@{}}
\toprule
Subspace & Basis from the SVD \\
\midrule
Row space \(C(A^{T})\) & \(v_1,\ldots,v_r\). \\
Nullspace \(N(A)\) & Remaining \(v_i\) with \(\sigma_i=0\). \\
Column space \(C(A)\) & \(u_1,\ldots,u_r\). \\
Left nullspace \(N(A^{T})\) & Remaining \(u_i\). \\
\bottomrule
\end{tabularx}
\end{center}

\section{SVD vs Eigenvalue Decomposition}

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.24\linewidth}Y Y@{}}
\toprule
Feature & Eigenvalue decomposition & SVD \\
\midrule
Basic form & \(A=S\Lambda S^{-1}\) & \(A=U\Sigma V^{T}\) \\
Works for & Diagonalizable square matrices & Every matrix \\
Numbers & Eigenvalues may be negative or complex & Singular values are nonnegative real numbers \\
Directions & Same space in and out & Input directions \(v_i\), output directions \(u_i\) \\
Best use & Dynamics and powers & Geometry, rank, least squares, approximation \\
\bottomrule
\end{tabularx}
\end{center}

\section{Symmetric Case}

For a symmetric positive definite matrix,
\[
A=Q\Lambda Q^{T}
\]
already has orthonormal eigenvectors and positive eigenvalues. In this special case, the SVD agrees with the eigenvalue decomposition.

For a symmetric matrix with negative eigenvalues, singular values use absolute sizes, not signs.

\section{Common Mistakes}

\begin{center}
\small
\begin{tabularx}{\linewidth}{@{}L{0.40\linewidth}Y@{}}
\toprule
Mistake & Correct exam habit \\
\midrule
Treating singular values as eigenvalues of \(A\). & Singular values come from \(A^{T}A\). \\
Allowing negative singular values. & Always take \(\sigma_i\ge 0\). \\
Using \(u_i=Av_i/\sigma_i\) when \(\sigma_i=0\). & Only divide by positive singular values. \\
Forgetting rectangular dimensions. & \(\Sigma\) has the same shape as \(A\). \\
Mixing up \(U\) and \(V\). & \(V\) belongs to the input side; \(U\) belongs to the output side. \\
\bottomrule
\end{tabularx}
\end{center}

\section{Final Exam Checklist}

\tightlabel{Checklist}
\begin{enumerate}
  \item Compute \(A^{T}A\).
  \item Find eigenvalues and orthonormal eigenvectors of \(A^{T}A\).
  \item Set \(\sigma_i=\sqrt{\lambda_i}\).
  \item Put the \(v_i\)'s into \(V\).
  \item Compute \(u_i=Av_i/\sigma_i\) for positive \(\sigma_i\).
  \item Complete orthonormal bases when needed.
  \item Write \(A=U\Sigma V^{T}\) with matching dimensions.
  \item Use zero singular values to identify nullspaces.
\end{enumerate}

\end{document}
